---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(desc)

#' Parse and format author list.
get_authors <- function() {
  authors <- format(eval(parse(text = desc_get("Authors@R"))),
                    include = c("given", "family"))
  paste0(authors, collapse = ", ")
}

#' Format description.
#'
#' @details Currently no formatting done, but could e.g., convert package names to hyperlinks.
get_description <- function() {
  desc_get("Description")
}

#' Create the CRAN installation instructions.
#'
#' @details If package is found on CRAN, provide [install.packages()] command and version number.  Otherwise do nothing.
cran_install <- function() {
  pkg_name <- desc_get("Package")
  # check if cran version exists
  cran_pkgs <- available.packages(filter = "CRAN")
  if(pkg_name %in% cran_pkgs[,"Package"]) {
    pkg_info <- cran_pkgs[pkg_name,]
    cat(
      paste0("To install the CRAN version (", pkg_info["Version"], "):"),
      "",
      "```r",
      paste0('install.packages("', pkg_name,
             '", INSTALL_opts = "--install-tests")'),
      "```",
      "",
      sep = "\n"
    )
  }
}

#' Create the GitHub install command.
github_install <- function() {
  pkg_name <- desc_get("Package")
  cat(
    "```r",
    paste0('devtools::install_github("mlysy/', pkg_name,
           '", INSTALL_opts = "--install-tests")'),
    "```",
    sep = "\n"
  )
}

#' Create call to package vignette.
get_vignette <- function() {
  pkg_name <- desc_get("Package")
  paste0('`vignette("', pkg_name, '")`')
}

#' Create call to run testthat.
get_tests <- function() {
  pkg_name <- desc_get("Package")
  cat(
    "```r",
    paste0('testthat::test_package("', pkg_name,
           '", reporter = "progress")'),
    "```",
    sep = "\n"
  )
}
```

# `r desc_get("Package")`: `r desc_get("Title")`

*`r get_authors()`*

<!-- badges: start -->
<!-- badges: end -->

---

### Description

`r get_description()`


### Installation

```{r cran-install, echo = FALSE, results = "asis"}
cran_install()
```

To install the latest development version: first install the [**devtools**](https://CRAN.R-project.org/package=devtools) package, then:

```{r github-install, echo = FALSE, results = "asis"}
github_install()
```

### Usage

A quick tutorial is provided in the package vignette: `r get_vignette()`.

### Unit Tests

To verify that the package has been installed correctly, you can run its unit tests.  First install the [**testthat**](https://CRAN.R-project.org/package=testthat) package, then:

```{r unit-tests, echo = FALSE, results = "asis"}
get_tests()
```
